Understanding the Socio-Economic Determinants of Voter Behavior: An Analysis of the 2020 U.S. Presidential Election

### 2024-03-13 paper/us_election_2020.qmd
new page for intro
fix citation brackets

# Introduction
Four years ago, the United States once again was at the forefront of an important electoral contest between the Democratic and Republican parties in the 2020 presidential election. The Democratic and Republican parties are represented by Donald Trump and Joe Biden respectively. Donald Trump achieved a remarkable victory in the 2016 election against Hillary Clinton. On the other hand, Joe Biden, the former vice-president under Barack Obama's administration from 2008 to 2016, led the Democratic campaign. He aimed to regain control from the Republican party. In the end, Biden won the election by 51.3% majority votes (@cnn_2020), which Trump questioned its credibility. Due to the COVID-19 pandemic in 2020, the vote system changed to in-mail voting instead of in-person voting, which was why Trump believed that some votes were stolen (@trump_stolenvote). Therefore, it is important for us to analyze why people voted for Biden and Trump in 2020 election and how various demographic, economic, and social factors may influence individuals' voting decisions. Perhaps then we could see why Biden eventually won. We analyze the 2020 election data set from Cooperative Election Study (CES) (@citeCes2020) and build a prediction model for individual voting preference. This allows us to discover trends and address potential political, economic, and societal situations that may cause the eventual winning for Biden.

@@

## Data Source
In this report, we use the 2020 Cooperative Election Study (CES) (@citeCes2020) as our primary dataset. This is a long-established annual political survey of US. CES contains information about how Americans view Congress and hold their representatives accountable in elections, how they vote and their electoral experiences, and how their behaviors and experiences vary with political geography and social context. In 2020, there were 61,000 American adults completed the survey.

@@

The cleaned data will be analyzed and performed using R (@citeR) with `tidyverse` (@citeTidyverse), `here` (@citeHere), `rstanarm` (@citeRstanarm), `modelsummary` (@citeModleSummary), `ggplot2` (@citeGgplot2), `knitr` (@citeKnitr), `marginaleffects` (@citeMarginalEffects), `plotly` (@citePlotly), `tibble` (@citeTibble), `margins` (@citeMargins), `testthat` (@citetestthat) and `kableExtra` (@citeKableExtra). @tbl-clean-data shows a preview of the cleaned dataset.

@@

The reason for the majority of Hispanics supporting Biden may be due to the US-Mexico border wall that was built during Trump's reign. The wall was built over 452 miles, which replaced the old fence (@Rodriguez_2021). With this measure in place, Trump lost most of the Hispanic vote. Trump claims the wall reformed the immigration system and achieved the most secure southern border in US history. However, that wasn't true, and with it has come a surge in illegal border crossings. For Hispanics, Trump's long history of anti-immigrant rhetoric and this border wall have denied them the opportunity to seek asylum. While crossing the border illegally is not something to be promoted, it is the last hope for those living in dire straits.

2020 is the age of pandemics, as Trump's phrase "Chinese virus" led to a large number of Asians starting to support Biden. This is an extremely anti-Asian statement: although COVID-19 was first found in China, but it does not mean that virus was from China. Trump's comments quickly festered in public opinion and led to a dramatic increase in hate incidents in the Asian American community (@Reja_2021). The repeated use of the "Chinese Virus" by a president of the country has contributed to an environment of racial discrimination and hatred against Asians, which has directly led to more Asians supporting Biden, whose policies are more open and democratic, and who values racial equality more.

@@

## Posterior predictive check
In @fig-post_dist, we implement a posterior distribution. This compares who people voted for in reality with the prediction results from the posterior distribution from our logistic regression model. It can be observed that the posterior distribution fits perfectly with the actual data, it suggests that the model accurately captures the observed data patterns. So the posterior is able to generate simulated data that closely resembles the actual data (@bayes_post). This is a positive sign because it indicates that our logistic regression model is a good representation of the actual voting preferences in the 2020 election data.

>>>>

# Introduction
Four years ago, the United States once again was at the forefront of an important electoral contest between the Democratic and Republican parties in the 2020 presidential election. The Democratic and Republican parties are represented by Donald Trump and Joe Biden respectively. Donald Trump achieved a remarkable victory in the 2016 election against Hillary Clinton. On the other hand, Joe Biden, the former vice-president under Barack Obama's administration from 2008 to 2016, led the Democratic campaign. He aimed to regain control from the Republican party. In the end, Biden won the election by 51.3% majority votes [@cnn_2020], which Trump questioned its credibility. Due to the COVID-19 pandemic in 2020, the vote system changed to in-mail voting instead of in-person voting, which was why Trump believed that some votes were stolen [@trump_stolenvote]. Therefore, it is important for us to analyze why people voted for Biden and Trump in 2020 election and how various demographic, economic, and social factors may influence individuals' voting decisions. Perhaps then we could see why Biden eventually won. We analyze the 2020 election data set from Cooperative Election Study (CES) [@citeCes2020] and build a prediction model for individual voting preference. This allows us to discover trends and address potential political, economic, and societal situations that may cause the eventual winning for Biden.

@@

## Data Source
In this report, we use the 2020 Cooperative Election Study (CES) [@citeCes2020] as our primary dataset. This is a long-established annual political survey of US. CES contains information about how Americans view Congress and hold their representatives accountable in elections, how they vote and their electoral experiences, and how their behaviors and experiences vary with political geography and social context. In 2020, there were 61,000 American adults completed the survey.

@@

The cleaned data will be analyzed and performed using R [@citeR] with `tidyverse` [@citeTidyverse], `here` [@citeHere], `rstanarm` [@citeRstanarm], `modelsummary` [@citeModleSummary], `ggplot2` [@citeGgplot2], `knitr` [@citeKnitr], `marginaleffects` [@citeMarginalEffects], `plotly` [@citePlotly], `tibble` [@citeTibble], `margins` [@citeMargins], `testthat` [@citetestthat] and `kableExtra` [@citeKableExtra]. @tbl-clean-data shows a preview of the cleaned dataset.

@@

The reason for the majority of Hispanics supporting Biden may be due to the US-Mexico border wall that was built during Trump's reign. The wall was built over 452 miles, which replaced the old fence [@Rodriguez_2021]. With this measure in place, Trump lost most of the Hispanic vote. Trump claims the wall reformed the immigration system and achieved the most secure southern border in US history. However, that wasn't true, and with it has come a surge in illegal border crossings. For Hispanics, Trump's long history of anti-immigrant rhetoric and this border wall have denied them the opportunity to seek asylum. While crossing the border illegally is not something to be promoted, it is the last hope for those living in dire straits.

2020 is the age of pandemics, as Trump's phrase "Chinese virus" led to a large number of Asians starting to support Biden. This is an extremely anti-Asian statement: although COVID-19 was first found in China, but it does not mean that virus was from China. Trump's comments quickly festered in public opinion and led to a dramatic increase in hate incidents in the Asian American community [@Reja_2021]. The repeated use of the "Chinese Virus" by a president of the country has contributed to an environment of racial discrimination and hatred against Asians, which has directly led to more Asians supporting Biden, whose policies are more open and democratic, and who values racial equality more.

@@

## Posterior predictive check
In @fig-post_dist, we implement a posterior distribution. This compares who people voted for in reality with the prediction results from the posterior distribution from our logistic regression model. It can be observed that the posterior distribution fits perfectly with the actual data, it suggests that the model accurately captures the observed data patterns. So the posterior is able to generate simulated data that closely resembles the actual data [@bayes_post]. This is a positive sign because it indicates that our logistic regression model is a good representation of the actual voting preferences in the 2020 election data.
